Effects of Different Pain Relievers and Dosages on Cognitive Retention

### 2025-04-06 paper/paper.qmd
File discussion done new updates

# Introduction

@@

For each treatment, we selected 30 different people (n = 30).
Through randomly picking the virtual citizens, we **replicated** 30 experimental units per treatment.
The even observations per treatment group ensures we have a balanced design.

@@

1.  Quantitative variables: Memory Game Scores, Memory Card Scores
2.  Qualitative variables: Type of pain reliever, dosage level (low/high)

@@

@tbl-summary-1 shows the summary statistics of the post-treatment memory game scores and memory card scores. The mean memory game score is 62.58 seconds, with a median of 61.50 seconds, which indicates that the distribution is fairly symmetric. The standard deviation (SD) is relatively high, indicating there is variability in overall participant performance. Looking at the  IQR (Interquartile Range), it is smaller than the SD, which suggests that most scores fall within a moderate range,  but there are some values  likely further from the center. For the memory card scores, the mean is 7.95/10 and the median is 9.00/10, which indicates a slight left skew in the distribution. Overall, the SD and IQR are both low, suggesting limited variability in these scores. While the summary statistics highlight some variability, they don’t directly indicate the presence of outliers; further visual analysis is required to assess this.

@@

# Compute statistics for Memory Game Score
mean_game <- mean(clean_data$memory_game_score, na.rm = TRUE)
median_game <- median(clean_data$memory_game_score, na.rm = TRUE)
sd_game <- sd(clean_data$memory_game_score, na.rm = TRUE)
iqr_game <- IQR(clean_data$memory_game_score, na.rm = TRUE)

# Compute statistics for Memory Cards Score
mean_cards <- mean(clean_data$memory_cards_score, na.rm = TRUE)
median_cards <- median(clean_data$memory_cards_score, na.rm = TRUE)
sd_cards <- sd(clean_data$memory_cards_score, na.rm = TRUE)
iqr_cards <- IQR(clean_data$memory_cards_score, na.rm = TRUE)

# Combine into a summary table
summary_table <- data.frame(
  Statistic = c("Mean", "Median", "SD", "IQR"),
  `Memory Game Score` = c(mean_game, median_game, sd_game, iqr_game),
  `Memory Cards Score` = c(mean_cards, median_cards, sd_cards, iqr_cards)
)

colnames(summary_table) <- c("Statistic", "Memory Game Score", "Memory Cards Score")

# Display table

kable(summary_table, format = "latex", booktabs = TRUE)

@@

p1 <- ggplot(clean_data, aes(x = drug, y = memory_game_score, color = dosage_level, group = dosage_level)) +
  stat_summary(fun = mean, geom = "point", size = 2) +
  stat_summary(fun = mean, geom = "line", size = 1) +
  labs(title = "Memory Game Scores",
       x = "Drug Type",
       y = "Mean Memory Game Score",
       color = "Dosage Level") +
  theme_minimal()

```
```{r}
#| label: fig-combined
#| echo: false
#| warning: false
#| message: false
#| fig.width: 12
#| fig.height: 6
#| fig-cap: "Interaction Plot and Boxplot: Memory Game Scores by Drug Type and Dosage Level"

@@

1. The median scores for all drugs appear similar before treatment.
2. Tramadol has a higher median than other drugs both before and after treatment.
3. The Placebo group has a slightly lower median score compared to other drugs.
4. There is one outlier in the Tramadol group before treatment.
5. The (IQR) is similar across all drug types, though Tramadol sees a decrease in its
IQR in the memory game scores after treatment.
These boxplots provide an initial comparison of cognitive performance changes before and after
drug administration.

@@

Our second assumption of *homogeneity of variance* is verified by the p-value = 0.3346 from the Bartlett test in @tbl-bartlett, which is greater than 0.05.
Due to this, we fail to reject the null hypothesis.
Lastly, for our two-way ANOVA model, the different combinations of pain reliever type and dosage levels, the treatments were randomly assigned to all participants which ensures *independence*.

@@

@@

-   $Y_{ijk}$: Memory score for the $k$-th participant receiving the $i$-th drug and $j$-th dosage level\
-   $\mu$: Overall mean\
-   $\alpha_i$: Effect of the $i$-th drug ((Aspirin, Tramadol, Paracetamol, Placebo))\
-   $\beta_j$: Effect of the $j$-th dosage level (High Dosage, Low dosage)\
-   $(\alpha\beta)_{ij}$: Interaction effect between drug and dosage\
-   $\epsilon_{ijk} \sim N(0, \sigma^2)$ : Random error

@@

The @tbl-model-result show the intercept ($\beta_0$ = 60.02) is the average memory game score for individuals given High dosage and Aspirin Drug (reference group). While comparing other treatments to reference level, we observe most coefficients from @tbl-model-result  are not statistically significant at 5% level. The coefficient of Drug Paracetamol (p-value = 0.741), Tramadol (p-value = 0.528), Placebo (p-value = 0.448) indicates that mean memory game scores for these groups are not statistically different from the reference group. Similarly, the evidence for all the Interaction terms are suggest no statistical significance between drug type and dosage level has any effect on the memory game score performance as p value are greater than 5% significance level. The lack of significance indicates that the effect of dosage does not vary meaningfully across drug types and dosage levels, indicating an absence of interaction between the two factors.

@@

Since no interaction effect was present, we examined the main effects independently. As mentioned, drug type was not significant, and neither was **dosage level** (\(p = 0.26\)), indicating that neither factor significantly influenced cognitive performance.

@@

Although the ANOVA model yielded no statistically significant results, we also considered the practical significance of the findings. The model’s $R^2 = 0.0398$ suggests that approximately 4\% of the variability in memory performance was accounted for by the treatment factors. However, in the context of ANOVA, $R^2$ is not typically used for interpretation, as it does not directly reflect differences between group means.

Instead, we refer to Cohen’s $f$ as a more appropriate measure of effect size in ANOVA. Cohen’s $f$ is based on the proportion of variance explained and provides context for the magnitude of the observed effects. [[[SHOULD WE STILL INCLUDE THIS?: For this study, we assume a medium effect size, corresponding to Cohen’s $f \approx 0.25$, based on conventional benchmarks from lecture materials. Even though the statistical test did not yield significant results. This suggests a small-medium effect size, indicating that the type of drug and dosage level may still have had a meaningful impact on memory task performance.]]]

@@

```{r}
# Compute statistics for Memory Game Score
mean_game <- mean(clean_data$memory_game_score, na.rm = TRUE)
median_game <- median(clean_data$memory_game_score, na.rm = TRUE)
sd_game <- sd(clean_data$memory_game_score, na.rm = TRUE)
iqr_game <- IQR(clean_data$memory_game_score, na.rm = TRUE)

# Compute statistics for Memory Cards Score
mean_cards <- mean(clean_data$memory_cards_score, na.rm = TRUE)
median_cards <- median(clean_data$memory_cards_score, na.rm = TRUE)
sd_cards <- sd(clean_data$memory_cards_score, na.rm = TRUE)
iqr_cards <- IQR(clean_data$memory_cards_score, na.rm = TRUE)

# Combine into a summary table
summary_table <- data.frame(
  Statistic = c("Mean", "Median", "SD", "IQR"),
  `Memory Game Score` = c(mean_game, median_game, sd_game, iqr_game),
  `Memory Cards Score` = c(mean_cards, median_cards, sd_cards, iqr_cards)
)

colnames(summary_table) <- c("Statistic", "Memory Game Score", "Memory Cards Score")

# Display table

kable(summary_table, format = "latex", booktabs = TRUE)
```

@@

# summary dataframe
age_summary <- data.frame(
  `Age Group` = names(age_counts),
  Count = as.numeric(age_counts),
  Percentage = round(as.numeric(age_percentages), 2)
)
colnames(age_summary)[colnames(age_summary) == "Age.Group"] <- "Age Group"

p1 <- ggplot(clean_data, aes(x = drug, y = memory_game_score, color = dosage_level, group = dosage_level)) +
  stat_summary(fun = mean, geom = "point", size = 3) +
  stat_summary(fun = mean, geom = "line", size = 1) +
  labs(title = "Memory Game Scores",
       x = "Drug Type",
       y = "Mean Memory Game Score",
       color = "Dosage Level") +
  theme_minimal()

# Interaction plot for Memory Cards Scores
p2 <- ggplot(clean_data, aes(x = drug, y = memory_cards_score, color = dosage_level, group = dosage_level)) +
  stat_summary(fun = mean, geom = "point", size = 3) +
  stat_summary(fun = mean, geom = "line", size = 1) +
  labs(title = "Memory Cards Scores",
       x = "Drug Type",
       y = "Mean Memory Cards Score",
       color = "Dosage Level") +
  theme_minimal()

# Combine the two plots vertically
combined_plot <- p1 / p2
combined_plot
```

>>>>

# Introduction {#sec-intro}

@@

For each treatment, we selected 30 different people (n = 30).
Through randomly picking the virtual citizens, we **replicated** 30 experimental units per treatment.
The even observations per treatment group ensures we have a **balanced design**.

@@

1.  Quantitative variables: Memory Game Scores
2.  Qualitative variables: Type of pain reliever, dosage level (low/high)

@@

@tbl-summary-1 shows the summary statistics of the treatment memory game scores. The mean memory game score is 62.58 seconds, with a median of 61.50 seconds, which indicates that the distribution is fairly symmetric.

@@

mean_game <- mean(clean_data$memory_game_score, na.rm = TRUE)
median_game <- median(clean_data$memory_game_score, na.rm = TRUE)
sd_game <- sd(clean_data$memory_game_score, na.rm = TRUE)
iqr_game <- IQR(clean_data$memory_game_score, na.rm = TRUE)

# Create summary table with clean column header
summary_table <- data.frame(
  Statistic = c("Mean", "Median", "SD", "IQR"),
  "Memory Game Score" = c(mean_game, median_game, sd_game, iqr_game),  check.names = FALSE
)

# Display table
kable(summary_table, format = "latex", booktabs = TRUE)

@@

p1 <- ggplot(clean_data, aes(x = drug, y = memory_game_score, color = dosage_level, group = dosage_level)) +
  stat_summary(fun = mean, geom = "point", size = 2) +
  stat_summary(fun = mean, geom = "line", size = 1) +
  labs(title = "Memory Game Scores",
       x = "Drug Type",
       y = "Mean Memory Game Score",
       color = "Dosage Level") +
  theme_minimal()
```

```{r}
#| label: fig-combined
#| echo: false
#| warning: false
#| message: false
#| fig.width: 12
#| fig.height: 6
#| fig-cap: "Interaction and Boxplots for Drug and Dosage Effects"

@@

1. The median memory game scores are relatively similar across most drug types.
2. Tramadol (high dosage) shows a higher median score compared to other treatments.
3. The Placebo group has a slightly lower median score than the others.
4. The interquartile range (IQR) is consistent across all groups, although Tramadol (high dosage) has slightly more variability.

@@

Our second assumption of Homogeneity of Variance appears to be satisfied. As shown in the @fig-residual-plot, the spread of the residual remains relatively consistent across the of fitted values. Further the result of the Bartlett test for homogeneity of variances from @tbl-bartlett, suggests the p-value is 0.3346. Since this p-value is greater than the 0.05 significance level, we fail to reject the null hypothesis of equal variances. Hence, the assumption is verified. 
Lastly, for our two-way ANOVA model, the different combinations of pain reliever type and dosage levels, the treatments were randomly assigned to all participants which ensures *independence*.

@@

```{r}
#| label: fig-residual-plot
#| echo: false
#| warning: false
#| message: false
#| fig.width: 9  
#| fig.height: 8  
#| dpi: 300
#| out.width: 60%
#| out.height: 50%
#| fig.align: "center"
#| fig-cap: "Residual plot: Check for Homoscedasticity"

plot(anova_model, which = 3, main = "", sub = "")
```

@@

-   $Y_{ijk}$: Memory score for the $k$-th participant receiving the $i$-th drug and $j$-th dosage level\
-   $\mu$: Overall mean\
-   $\alpha_i$: Effect of the $i$-th drug (Aspirin, Tramadol, Paracetamol, Placebo)
-   $\beta_j$: Effect of the $j$-th dosage level (High Dosage, Low dosage)\
-   $(\alpha\beta)_{ij}$: Interaction effect between drug and dosage\
-   $\epsilon_{ijk} \sim N(0, \sigma^2)$ : Random error

@@

The @tbl-model-result show the intercept ($\beta_0$ = 62.20) is the average memory game score for individuals given High dosage and Aspirin Drug (reference group). While comparing other treatments to reference level, we observe most coefficients from @tbl-model-result  are not statistically significant at 5% level. The coefficient of Drug Paracetamol ($p-value = 0.9927$), Tramadol ($p-value = 0.5281$), Placebo ($p-value = 0.3893$) indicates that mean memory game scores for these groups are not statistically different from the reference group. Similarly, the evidence for all the Interaction terms suggests no statistical significance between drug type and dosage level has any effect on the memory game score performance as p value are greater than 5% significance level. The lack of significance indicates that the effect of dosage does not vary meaningfully across drug types and dosage levels, indicating an absence of interaction between the two factors.

@@

Since no interaction effect was present, we examined the main effects independently. As mentioned, drug type was not significant, and neither was **dosage level** ($p = 0.26$), indicating that neither factor significantly influenced cognitive performance.

@@

Since the interaction effect was not statistically significant ($p > 0.05$), we proceeded to assess the effect size with the purpose to evaluate the practical significance of our findings. Our model yields  $R^2 = 0.0398$. While $R^2$ commonly used in regression analysis to a measure explained variance, determining the goodness of fit of a model. However, in ANOVA Model the $R^2$ value doesn't have a direct interpretation as the proportion of variance explained by regression model and it also does not directly reflect differences between group means. We use the $R^2$ value to report the effect that corresponds to Cohen’s f guidelines i.e. $| f |$ is approximately 0.25 as our  $R^2$ falls between the small-medium categories.Even though the statistical test did not yield significant results, this quantity suggests we have a small-medium effect size, indicating that the type of drug and dosage level may still have had a meaningful impact on memory task performance.

@@

```{r}
mean_game <- mean(clean_data$memory_game_score, na.rm = TRUE)
median_game <- median(clean_data$memory_game_score, na.rm = TRUE)
sd_game <- sd(clean_data$memory_game_score, na.rm = TRUE)
iqr_game <- IQR(clean_data$memory_game_score, na.rm = TRUE)

# Create summary table with clean column header
summary_table <- data.frame(
  Statistic = c("Mean", "Median", "SD", "IQR"),
  "Memory Game Score" = c(mean_game, median_game, sd_game, iqr_game),  check.names = FALSE
)

# Display table
kable(summary_table, format = "latex", booktabs = TRUE)

@@

# Display table
kable(drug_summary, format = "latex", booktabs = TRUE)
```

```{r}
p3 <- ggplot(clean_data, aes(x = drug, y = memory_game_score, fill = dosage_level)) +
  geom_boxplot() +
  labs(x = "Drug Type", y = "Memory Game Score", fill = "Dosage Level") +
  theme_minimal()
```

```{r}
p1 <- ggplot(clean_data, aes(x = drug, y = memory_game_score, color = dosage_level, group = dosage_level)) +
  stat_summary(fun = mean, geom = "point", size = 2) +
  stat_summary(fun = mean, geom = "line", size = 1) +
  labs(title = "Memory Game Scores",
       x = "Drug Type",
       y = "Mean Memory Game Score",
       color = "Dosage Level") +
  theme_minimal()
```

```{r}
# Combine the plots side by side
p1 + p3
```
